---
title: "Cedars GIS project"
output: 
  md: "README.md"
editor: visual
---

```{r}
#| echo: false
```


## Load all the required libraries

```{r}
library(sf)
library(tidyverse)
library(rosm)
library(ggspatial)
library(raster)
library(leaflet)
library(htmltools)
library(rinat)
library(rgbif)
```

## Read in the data

For each data source:

1.  Read in the data

-   Widdringtonia cedarbergensis data from 2013 Google Earth Satellite Imagery from Slingsby & Slingsby (2019)
-   Field Widdringtonia cedarbergensis data from Slingsby & Slingsby (2019)
-   Repeat photography data from White et al. (2016)

2.  Check the CRS of the data
3.  Convert data to spatial framework if required

```{r}
# read in cedar trees mapped from 2013 satellite imagery
ced <- read_sf("data/All_Cedars_2_WGS84.kml")
head(ced)
class(ced)
st_crs(ced)

# read in cedar trees from Slingsby field survey 
field_ced <- read_sf("data/fieldtrees.gpkg")
st_crs(field_ced)

# read in repear photography data
photo_ced <- read_sf("data/data_raw_repeat_photography_White-2016.csv")
photo_ced <- st_as_sf(photo_ced, coords = c("Longitude", "Latitude"), crs = st_crs(ced))
st_crs(photo_ced)

# read in inat data for clanwilliam cedars
inat_ced <- get_inat_obs(taxon_name = "Widdringtonia cedarbergensis")

inat_ced <- inat_ced %>% 
  filter(positional_accuracy<46 & 
           latitude<0 & # remove points which were entered incorrectly as negative
           !is.na(latitude) &  # remove na values
           !is.na(longitude) &
           captive_cultivated == "false" & 
           quality_grade == "research")

inat_ced <- st_as_sf(inat_ced, coords = c("longitude", "latitude"), crs = 4326) # convert to spatial framework

# read in gbif data for all clanwilliam cedars 
gbif_ced <- (occ_data(scientificName = "Widdringtonia cedarbergensis",
                      hasCoordinate=TRUE, 
                      hasGeospatialIssue=FALSE))$data %>% 
  filter(coordinateUncertaintyInMeters<50 & 
           decimalLatitude<0 &
           !is.na("decimalLatitude"))

gbif_ced <- st_as_sf(gbif_ced, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

# read in gbif data for all Widdringtonia species 
gbif_all <- (occ_data(scientificName = "Widdringtonia",
                      hasCoordinate=TRUE, 
                      hasGeospatialIssue=FALSE))$data %>% 
  filter(coordinateUncertaintyInMeters<50 & 
           decimalLatitude<0 &
           !is.na("decimalLatitude"))

gbif_all <- st_as_sf(gbif_all, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
unique(gbif_all$scientificName)

```
## Plot the data

Create interactive maps of the Widdringtonia species data

The first map shows the distribution of three Widdringtonia species, W. cedarbergensis, W. nodiflora and W. schwarzii, across Southern Africa with data from GBIF.

```{r}
leaflet() %>%
  addTiles(group = "Default") %>%  
  addCircleMarkers(data = subset(gbif_all, scientificName == "Widdringtonia cedarbergensis J.A.Marsh"),
                   radius = 3, 
                   color = "green",
                   popup = ~paste("Scientific Name: ", scientificName, "<br>",
                                  "Date of record: ", year)) %>% 
  addCircleMarkers(data = subset(gbif_all, scientificName == "Widdringtonia nodiflora (L.) Powrie"),
                   radius = 3, 
                   color = "blue",
                   popup = ~paste("Scientific Name: ", scientificName, "<br>",
                                  "Date of record: ", year)) %>% 
  addCircleMarkers(data = subset(gbif_all, scientificName == "Widdringtonia schwarzii (Marloth) Mast."),
                   radius = 3, 
                   color = "red",
                   popup = ~paste("Scientific Name: ", scientificName, "<br>",
                                  "Date of record: ", year)) %>%
addLegend(position = "topright", 
            colors = c("green", "blue", "red"), 
            labels = c("Widdringtonia cedarbergensis", 
                       "Widdringtonia nodiflora", 
                       "Widdringtonia schwarzii"),
            title = "Species Key")
```
Click on an observation to view the species and the date the record was imported into GBIF.

The second map analyses the observations of W. cedarbergensis from iNaturalist and GBIF compared to the cedars mapped from the 2013 satellite imagery by Slingsby & Slingsby (2019).

```{r}
# interactive plot of satellite cedars, inat clanwilliam cedars and gbif clanwilliam cedars
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add points
  addCircleMarkers(data = ced,
                   radius = 3, 
                   color = "green") %>%
  addCircleMarkers(data = gbif_ced,
                   radius = 3,
                   color = "blue") %>%
  addCircleMarkers(data = inat_ced,
                   radius = 3,
                   color = "red") %>%
addLegend(position = "topright", 
            colors = c("green", "blue", "red"), 
            labels = c("Satellite observations", 
                       "GBIF observations", 
                       "iNaturalist observations"),
            title = "Observation Key")

```