---
title: "Cedars GIS project"
output: html
editor: visual
---

```{r}
#| echo: false
```

## Load the required libraries

```{r}
library(sf)
library(tidyverse)
library(rosm)
library(ggspatial)
library(leaflet)
library(htmltools)
library(rinat)
library(rgbif)
library(prettymapr)
```

## Read in the data

For each data source:

1.  Read in the data and filter if required
2.  Check the CRS of the data
3.  Convert data to spatial framework if required

***Widdringtonia cedarbergensis*** **data from 2013 Google Earth Satellite Imagery (Slingsby & Slingsby, 2019)**

Dataset license: CC BY 4.0

```{r}
# read in data
sat_ced <- read_sf("data/All_Cedars_2_WGS84.kml")
# check data class
class(sat_ced)
# check CRS
st_crs(sat_ced)
```

**Field *Widdringtonia cedarbergensis* data (Slingsby & Slingsby, 2019)**

Dataset license: CC BY 4.0

```{r}
# read in data
field_ced <- read_sf("data/fieldtrees.gpkg")
st_crs(field_ced)
```

**Repeat photography *Widdringtonia cedarbergensis* data (White et al., 2016)**

Dataset license: CC BY 3.0

```{r}
# read in data
photo_ced <- read_sf("data/data_raw_repeat_photography_White-2016.csv")
# convert to spatial framework
photo_ced <- st_as_sf(photo_ced, coords = c("Longitude", "Latitude"), crs = 4326)
```

**iNaturalist data for *Widdringtonia cedarbergensis***

```{r}
# read in data
inat_ced <- get_inat_obs(taxon_name = "Widdringtonia cedarbergensis")
# filter data
inat_ced <- inat_ced %>% 
  filter(positional_accuracy<46 & 
           latitude<0 & # remove points which were entered incorrectly as negative
           !is.na(latitude) &  # remove na values
           !is.na(longitude) &
           captive_cultivated == "false" & 
           quality_grade == "research") # remove captive or cultivated plants
# convert to spatial framework
inat_ced <- st_as_sf(inat_ced, coords = c("longitude", "latitude"), crs = 4326) 
```

**iNaturalist data for *Widdringtonia nodiflora***

```{r}
# read in data
inat_nod <- get_inat_obs(taxon_name = "Widdringtonia nodiflora")
# filter data
inat_nod <- inat_nod %>% 
  filter(positional_accuracy<46 & 
           latitude<0 & # remove points which were entered incorrectly as negative
           !is.na(latitude) &  # remove na values
           !is.na(longitude) &
           captive_cultivated == "false" & 
           quality_grade == "research") # remove captive or cultivated plants
# convert to spatial framework
inat_nod <- st_as_sf(inat_nod, coords = c("longitude", "latitude"), crs = 4326)
```

**iNaturalist data for *Widdringtonia whytei***

```{r}
# read in data
inat_sch <- get_inat_obs(taxon_name = "Widdringtonia schwarzii")
# filter data
inat_sch <- inat_sch %>% 
  filter(positional_accuracy<46 & 
           latitude<0 & # remove points which were entered incorrectly as negative
           !is.na(latitude) &  # remove na values
           !is.na(longitude) &
           captive_cultivated == "false" & 
           quality_grade == "research") # remove captive or cultivated plants
# convert to spatial framework
inat_sch <- st_as_sf(inat_sch, coords = c("longitude", "latitude"), crs = 4326)
```

**iNaturalist data for *Widdringtonia whytei***

```{r}
# read in data
inat_why <- get_inat_obs(taxon_name = "Widdringtonia whytei")
# filter data
inat_why <- inat_why %>% 
  filter(positional_accuracy<46 & 
           latitude<0 & # remove points which were entered incorrectly as negative
           !is.na(latitude) &  # remove na values
           !is.na(longitude) &
           captive_cultivated == "false" & 
           quality_grade == "research") # remove captive or cultivated plants
# convert to spatial framework
inat_why <- st_as_sf(inat_why, coords = c("longitude", "latitude"), crs = 4326)
```

**Global Biodiversity Information Facility (GBIF) data for *Widdringtonia cedarbergensis***

```{r}
# read in GBIF data for Widdringtonia cedarbergensis and filter data
gbif_ced <- (occ_data(scientificName = "Widdringtonia cedarbergensis",
                      hasCoordinate=TRUE, 
                      hasGeospatialIssue=FALSE))$data %>% 
  filter(coordinateUncertaintyInMeters<50 & 
           decimalLatitude<0 &
           !is.na("decimalLatitude"))
# convert to spatial framework
gbif_ced <- st_as_sf(gbif_ced, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
```

## Plot interactive maps

The first map we will plot shows the distribution of the four *Widdringtonia* species, *W. cedarbergensis, W. nodiflora,*, *W. schwarzii,* and *W. whytei* across Southern Africa from iNaturalist data.

```{r, }
#| fig-cap: "Interactive map of Widdringtonia species from iNaturalist"
#| label: fig-1

# plot interactive map using leaflet
leaflet() %>%
  addTiles(group = "Default") %>%  
  addCircleMarkers(data = inat_ced,
                   radius = 3, 
                   color = "green",
                   # add popup when you click on point
                   popup = ~paste("Scientific Name: ", scientific_name, "<br>",
                                  "Date of record: ", datetime)) %>% 
  addCircleMarkers(data = inat_nod,
                   radius = 3, 
                   color = "blue",
                   popup = ~paste("Scientific Name: ", scientific_name, "<br>",
                                  "Date of record: ", datetime)) %>% 
  addCircleMarkers(data = inat_sch,
                   radius = 3, 
                   color = "cyan",
                   popup = ~paste("Scientific Name: ", scientific_name, "<br>",
                                  "Date of record: ", datetime)) %>%
  addCircleMarkers(data = inat_why,
                   radius = 3, 
                   color = "darkblue",
                   popup = ~paste("Scientific Name: ", scientific_name, "<br>",
                                  "Date of record: ", datetime)) %>%
  # add a legend to the plot
  addLegend(position = "topright", 
            colors = c("green", "blue", "cyan", "darkblue"), 
            labels = c("Widdringtonia cedarbergensis",
                       "Widdringtonia nodiflora", 
                       "Widdringtonia schwarzii",
                       "Widdringtonia whytei"),
            title = "Species Key")
```

::: {.callout-tip appearance="simple"}
Click on an observation to view the species and the date the tree was observed.
:::

This map can be used to interrogate the distribution of the four *Widdringtonia* species, a genus native to Southern Africa. The map shows that *W. cedarbergensis* is confined to the Cederberg mountains in South Africa, *W. schwarzii* is confined to the Baviaanskloof and Kouga mountain ranges in the Eastern Cape, while *W. whytei* is endemic to Malawi. *Widdringtonia nodiflora* has a more widespread distribution across Southern Africa from the Western Cape to Malawi. The distribution of the four *Widdringtonia* species raises interesting questions about the evolution of the genus.

::: {.callout-note appearance="simple"}
Observations of *W. schwarzii* occur in Kirstenbosch National Botanical Garden, outside of their natural range, indicating these individuals have not been filtered out from the iNaturalist data set.
:::

The second map analyses the observations of *W. cedarbergensis* from iNaturalist and GBIF compared to the cedars mapped from the 2013 satellite imagery by Slingsby & Slingsby (2019).

```{r}
#| fig-cap: "Interactive map of Widdringtonia cedarbergensis from iNaturalist, GBIF, and 2013 satellite imagery."
#| label: fig-2

# plot interactive map using leaflet
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add points
  addCircleMarkers(data = sat_ced,
                   radius = 3, 
                   color = "green") %>%
  addCircleMarkers(data = gbif_ced,
                   radius = 3,
                   color = "blue",
                   popup = ~paste("Date of record: ", eventDate)) %>%
  addCircleMarkers(data = inat_ced,
                   radius = 3,
                   color = "cyan",
                   popup = ~paste("Date of record: ", datetime)) %>%
addLegend(position = "topright", 
            colors = c("green", "blue", "cyan"), 
            labels = c("Satellite observations", 
                       "GBIF observations", 
                       "iNaturalist observations"),
            title = "Observation Key")

```

::: {.callout-tip appearance="simple"}
Click on an observation to view the date the tree was observed.
:::

::: {.callout-note appearance="simple"}
You can see how iNaturalist and GBIF observations of cedars largely occur along hiking trails.
:::

The third map plots the in-field observations of *W. cedarbergensis* from Slingsby & Slingsby (2019) and the repeat photography *W. cedarbergensis* data from White et al. (2016), both of which are separated into alive and dead trees. The cedars mapped from the 2013 satellite imagery by Slingsby & Slingsby (2019) are also mapped.

```{r}
#| fig-cap: "Interactive map of Widdringtonia cedarbergensis from 2013 satellite imagery, in-field observations, and repeat photography observations. Trees observed in the field and by repeat photography have been coloured by alive and dead trees."
#| label: fig-3

# plot interactive map using leaflet
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add points
  addCircleMarkers(data = sat_ced,
                   radius = 3, 
                   color = "black") %>%
  addCircleMarkers(data = subset(photo_ced, Mortality == "1"),
                   radius = 3, 
                   color = "red",
                   popup = ~paste("Survey type: Repeat photography")) %>%
  addCircleMarkers(data = subset(photo_ced, Mortality == "0"),
                   radius = 3, 
                   color = "green",
                   popup = ~paste("Survey type: Repeat photography")) %>%
  addCircleMarkers(data = subset(field_ced, State == " Dead (Survey)"),
                   radius = 3, 
                   color = "red",
                   popup = ~paste("Survey type: In-field observations")) %>%
  addCircleMarkers(data = subset(field_ced, State == " Alive (Survey)"),
                   radius = 3, 
                   color = "green",
                   popup = ~paste("Survey type: In-field observations")) %>% 
addLegend(position = "topright", 
            colors = c("black", "green", "red"), 
            labels = c("Satellite tree observations", 
                       "Trees observed alive", 
                       "Trees observed dead"),
            title = "Key")
```

::: {.callout-tip appearance="simple"}
Click on an observation to view whether the tree observed was an in-field survey or repeat photography survey.
:::

# References

Slingsby, J.A. & Slingsby, P.W.O. 2019. Global distribution of Clanwilliam cedar tree localities in 2013. figshare \[dataset\] https://doi.org/10.6084/m9.figshare.7670435.v1

Slingsby, J.A. & Slingsby, P.W.O. 2019. Monitoring the critically endangered Clanwilliam cedar with freely available Google Earth imagery. PeerJ. https://doi.org/10.7717/peerj.7005

White, J.D.M., Jack, S., Hoffman, M.T., Puttick, J., Bonora, D., Visser, V., February, E. 2016. Collapse of an iconic conifer: long-term changes in the demography of Widdringtonia cedarbergensis using repeat photography. BMC Ecology, 16(1), 11 https://doi.org/10.1186/s12898-016-0108-6

White, J.D.M., Jack, S., Puttick, J., Bonora, D., Visser, V., February, E. 2016. Mortality and climatic/environmental correlates of Widdringtonia cedargensis trees in the Cederberg, South Africa. PANGAEA \[dataset\] https://doi.org/10.1594/PANGAEA.866937
